Unshuffling Permutations

### 2016-02-03 new-article.tex
LATIN 2016

\begin{lemma}
  \label{lemma:matching}
  Let $\pi$ be a permutation. The following statements are equivalent:
  \begin{enumerate}
    \item The permutation $\pi$ is a square.
    \item There exists an oriented perfect matching $\vv{\mathcal{M}}$
    of the linear graph $G_\pi$ that avoids
    \InclusionLL,~\InclusionLR,~\InclusionRL,~\InclusionRR,~\CrossingLR
    and \CrossingRL, and such that for any two distinct arcs $(a, a)$ and
    $(b, b')$ in $\vv{\mathcal{M}}$ we have $\pi(a) < \pi(b)$ if and only
    if $\pi(a') < \pi(b')$.
  \end{enumerate}
\end{lemma}

@@

    (\emph{2} $\Rightarrow$ \emph{1}).
    Let $I^1 = (i^1_1 < i^1_2 < \dots < i^1_n)$
    (resp. $I^2 = (i^2_1 < i^2_2 < \dots < i^2_n)$)
    be the ordered sequence of the origins (resp. sinks) of the arcs of
    the oriented perfect matching $\vv{\mathcal{M}}$.
    We first show that, for every $1 \leq j \leq n$,
    $(i^1_j, i^2_j)$ is an arc of $\vv{\mathcal{M}}$.
    The proof is by induction on $n$.
    Since the assertion is certainly valid for $n = 1$,
    it is enough to show that $(i^1_n, i^2_n)$ is an arc of $\vv{\mathcal{M}}$.
    Suppose, aiming at a contradiction that $(i^1_n, i^2_n)$
    is not an arc of $\vv{\mathcal{M}}$.
    Let $i^2_p$ be the sink of the arc with origin $i^1_n$ in
    $\vv{\mathcal{M}}$ (\emph{i.e.}, $(i^1_n, i^2_p)$ is an abstractof
    $\vv{\mathcal{M}}$).
    We need to consider two cases.
    \begin{itemize}
      \item If $i^1_n > i^2_p$ then there exist $i^1_q$ and $i^2_r$ such
      that
      \begin{center}
        \begin{tabular}{ccc}
          \LabeledInclusionRL{.3}{$i^2_p$}{$i^1_q$}{$i^2_r$}{$i^1_n$}
          &\qquad
          \LabeledInclusionRR{.3}{$i^1_p$}{$i^2_r$}{$i^1_q$}{$i^1_n$}
          &\qquad
          \LabeledCrossingLR{.3}{$i^1_p$}{$i^2_p$}{$i^2_r$}{$i^1_n$}
        \end{tabular}
      \end{center}
      This is a contradiction since
      $\vv{\mathcal{M}}$ avoids
      \InclusionLL,~\InclusionLR,~\InclusionRL,~\InclusionRR,~\CrossingLR and
      \CrossingRL.
      \item
      If $i^1_n < i^2_p$ then there exist $i^1_q$ and $i^2_r$ such
      that
      \begin{center}
        \begin{tabular}{ccc}
          \LabeledInclusionLL{.3}{$i^1_q$}{$i^1_n$}{$i^2_p$}{$i^2_r$}
          &\qquad
          \LabeledInclusionRL{.3}{$i^2_r$}{$i^1_n$}{$i^2_p$}{$i^1_q$}
          &\qquad
          \LabeledCrossingRL{.3}{$i^2_p$}{$i^1_q$}{$i^1_n$}{$i^2_r$}
        \end{tabular}
      \end{center}
      This is a contradiction since
      $\vv{\mathcal{M}}$ avoids
      \InclusionLL,~\InclusionLR,~\InclusionRL,~\InclusionRR,~\CrossingLR and
      \CrossingRL.
    \end{itemize}
    Then it follows that $(i^1_n, i^2_n)$ is an arc
    of $\vv{\mathcal{M}}$.
    Now, let $p^1 = \pi(i^1_1) \pi(i^1_2) \dots \pi(i^1_n)$ and
    $p^2 = \pi(j^2_1) \pi(^2_2) \dots \pi(i^2_n)$ be the two patterns
    from $\pi$ induced by $I^1$ and $I^2$, respectively.
    Clearly $p^1$ and $p^2$ are disjoint in $\pi$
    (since $\vv{\mathcal{M}}$ is a matching)
    and cover $\pi$ (since $\vv{\mathcal{M}}$ is perfect).
    We claim that $p^1$ and $p^2$ are order-isomorphic.
    Indeed, this is a direct consequence of th fact that,
    for any two distinct edges $(a, b)$ and $(a', b')$ in
    $\vv{\mathcal{M}}$ we have $\pi(a) < \pi(a')$ if and only if
    $\pi(b) < \pi(b')$.
    \qed
\end{proof}

@@

\begin{lemma}
  \label{lemma:at most one edge monotone}
  Let $\pi = \pi_1 \, \sigma_1 \, \pi_2 \, \sigma_2 \, \pi_3$
  be a permutation with $|\sigma_1| \geq 2$ and $|\sigma_2| \geq 2$,
  and $\mathcal{M}$ be a containment-free perfect matching of $G_\pi$.
  The following assertions hold:
  \begin{enumerate}
    \item
    If $\sigma_1$ is increasing, $\sigma_2$ is decreasing, and
    $\sigma_1$ is above $\sigma_2$ (see Fig.~\ref{subfig:increasing before
    and above decreasing}), then $\mathcal{M}$ contains at most one
    $(\sigma_1, \sigma_2)$-edge.
    \item
    If $\sigma_1$ is decreasing, $\sigma_2$ is increasing, and $\sigma_1$
    is below $\sigma_2$ (see Fig.~\ref{subfig:decreasing before and below
    increasing}), then $\mathcal{M}$ contains at most one
    $(\sigma_1, \sigma_2)$-edge.
  \end{enumerate}
\end{lemma}

\begin{proof}[of Lemma~\ref{lemma:at most one edge monotone}]
  \begin{figure}[t!]
    \centering
    \subfigure[%
      An increasing pattern before and above a decreasing pattern.
    ]{%
      \begin{tikzpicture}
        [
          scale=0.325,
          label/.style={anchor=base}
        ]
        % increasing pattern
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,5.8) grid (5.2,11.2);
        \foreach \x/\y in {0/6,1/7,2/8,3/9,4/10,5/11} {
            \draw [fill=black] (\x,\y) circle (0.1);
        }
        % decreasing pattern
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,-0.2) grid (11.2,5.2);
        \foreach \x/\y in {6/5,7/4,8/3,9/2,10/1,11/0} {
            \draw [fill=black] (\x,\y) circle (0.1);
        }
        % labels
        \node [label] (a) at (1,5) {$a$};
        \node [label] (b) at (4,5) {$b$};
        \node [label] (ap) at (7,-1) {$a'$};
        \node [label] (bp) at (10,-1) {$b'$};
        \node (nu1) at (2.5,4) {$\sigma_1$};
        \node (nu2) at (8.5,-2) {$\sigma_2$};
        % crossings edges
        \draw [line width=1pt]
        (1,7) .. controls +(0,12) and +(0,10) .. (7,4);
        \draw [line width=1pt]
        (4,10) .. controls +(0,12) and +(0,10) .. (10,1);
      \end{tikzpicture}
      \label{subfig:increasing before and above decreasing}
    }% end subfigure
    \qquad\qquad\qquad
    \subfigure[%
      A decreasing pattern before and below an increasing pattern.
    ]{%
    \begin{tikzpicture}
      [
        scale=0.325,
        label/.style={anchor=base}
      ]
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,-0.2) grid (5.2,5.2);
      \foreach \x/\y in {0/5,1/4,2/3,3/2,4/1,5/0} {
          \draw [fill=black] (\x,\y) circle (0.1);
      }
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,5.8) grid (11.2,11.2);
      \foreach \x/\y in {6/6,7/7,8/8,9/9,10/10,11/11} {
          \draw [fill=black] (\x,\y) circle (0.1);
      }
      % labels
      \node [label] (a) at (1,-1) {$a$};
      \node [label] (b) at (4,-1) {$b$};
      \node [label] (ap) at (7,5) {$a'$};
      \node [label] (bp) at (10,5) {$b'$};
      \node (nu1) at (2.5,-2) {$\sigma_1$};
      \node (nu2) at (8.5,4) {$\sigma_2$};
      % crossings edges
      \draw [line width=1pt]
      (1,4) .. controls +(0,10) and +(0,12) .. (7,7);
      \draw [line width=1pt]
      (4,1) .. controls +(0,10) and +(0,12) .. (10,10);
      \end{tikzpicture}
      \label{subfig:decreasing before and below increasing}
    }% end subfigure
    \caption{\label{fig:subfig:no (nu_1, nu_2)-edge}%
      Illustration of Lemma~\ref{lemma:at most one edge monotone}.
    }
  \end{figure}
  We first prove \emph{1}. Suppose, aiming at a contradiction, that
  there exist two $(\sigma_1, \sigma_2)$-edges $(a, a')$ and $(b, b')$
  in $\mathcal{M}$. Since $\mathcal{M}$ is containment-free, we may
  assume $a < b$ and $a' < b'$. Therefore, we have $\pi(a) < \pi(b)$
  (since $\sigma_1$ is increasing) and $\pi(a') > \pi(b')$ (since
  $\sigma_2$ is decreasing), which contradicts our hypothesis about
  $\mathcal{M}$.

  As for \emph{2}, suppose, still aiming at a contradiction, that there
  exist two $(\sigma_1, \sigma_2)$-edges $(a, a')$ and $(b, b')$ in
  $\mathcal{M}$. Since $\mathcal{M}$ is containment-free, we may assume
  $a < b$ and $a' < b'$. Therefore, we have $\pi(a) > \pi(b)$ (since
  $\sigma_1$ is decreasing) and $\pi(a') < \pi(b')$ (since $\sigma_2$ is
  increasing), which contradicts our hypothesis about~$\mathcal{M}$.
  \qed
\end{proof}

>>>>

\begin{lemma}
  \label{lemma:matching}
  Let $\pi$ be a permutation. The following statements are equivalent:
  \begin{enumerate}
    \item The permutation $\pi$ is a square.
    \item There exists an oriented perfect matching $\vv{\mathcal{M}}$
    of the linear graph $G_\pi$ that avoids
    \InclusionLL,~\InclusionLR,~\InclusionRL,~\InclusionRR,~\CrossingLR
    and \CrossingRL, and such that for any two distinct arcs $(a, a')$ and
    $(b, b')$ in $\vv{\mathcal{M}}$ we have $\pi(a) < \pi(b)$ if and only
    if $\pi(a') < \pi(b')$.
  \end{enumerate}
\end{lemma}

@@

    (\emph{2} $\Rightarrow$ \emph{1}).
    Let $I^1 = (i^1_1 < i^1_2 < \dots < i^1_n)$
    (resp. $I^2 = (i^2_1 < i^2_2 < \dots < i^2_n)$)
    be the ordered sequence of the origins (resp. sinks) of the arcs of
    the oriented perfect matching $\vv{\mathcal{M}}$.
    We first show that, for every $1 \leq j \leq n$,
    $(i^1_j, i^2_j)$ is an arc of $\vv{\mathcal{M}}$.
    The proof is by induction on $n$.
    Since the assertion is certainly valid for $n = 1$,
    it is enough to show that $(i^1_n, i^2_n)$ is an arc of $\vv{\mathcal{M}}$.
    Suppose, aiming at a contradiction that $(i^1_n, i^2_n)$
    is not an arc of $\vv{\mathcal{M}}$.
    Let $i^2_p$ be the sink of the arc with origin $i^1_n$ in
    $\vv{\mathcal{M}}$ (\emph{i.e.}, $(i^1_n, i^2_p)$ is an arc of
    $\vv{\mathcal{M}}$).
    We need to consider two cases.
    \begin{itemize}
      \item If $i^1_n > i^2_p$ then there exist $i^1_q$ and $i^2_r$ such
      that
      \begin{center}
        \begin{tabular}{cccc}
          \LabeledInclusionRL{.3}{$i^2_p$}{$i^1_q$}{$i^2_r$}{$i^1_n$}
          &\qquad
          \LabeledInclusionRR{.3}{$i^1_p$}{$i^2_r$}{$i^1_q$}{$i^1_n$}
          &\qquad
          \LabeledCrossingLR{.3}{$i^1_p$}{$i^2_p$}{$i^2_r$}{$i^1_n$}
          &\qquad
          \LabeledCrossingRL{.3}{$i^2_p$}{$i^1_q$}{$i^1_n$}{$i^2_r$}
        \end{tabular}
      \end{center}
      This is a contradiction since
      $\vv{\mathcal{M}}$ avoids
      \InclusionLL,~\InclusionLR,~\InclusionRL,~\InclusionRR,~\CrossingLR and
      \CrossingRL.
      \item
      If $i^1_n < i^2_p$ then there exist $i^1_q$ and $i^2_r$ such
      that
      \begin{center}
        \begin{tabular}{cc}
          \LabeledInclusionLL{.3}{$i^1_q$}{$i^1_n$}{$i^2_p$}{$i^2_r$}
          &\qquad
          \LabeledInclusionRL{.3}{$i^2_r$}{$i^1_n$}{$i^2_p$}{$i^1_q$}
        \end{tabular}
      \end{center}
      This is a contradiction since
      $\vv{\mathcal{M}}$ avoids
      \InclusionLL,~\InclusionLR,~\InclusionRL,~\InclusionRR,~\CrossingLR and
      \CrossingRL.
    \end{itemize}
    Then it follows that $(i^1_n, i^2_n)$ is an arc
    of $\vv{\mathcal{M}}$.
    Now, let $p^1 = \pi(i^1_1) \pi(i^1_2) \dots \pi(i^1_n)$ and
    $p^2 = \pi(j^2_1) \pi(^2_2) \dots \pi(i^2_n)$ be the two patterns
    from $\pi$ induced by $I^1$ and $I^2$, respectively.
    Clearly $p^1$ and $p^2$ are disjoint in $\pi$
    (since $\vv{\mathcal{M}}$ is a matching)
    and cover $\pi$ (since $\vv{\mathcal{M}}$ is perfect).
    We claim that $p^1$ and $p^2$ are order-isomorphic.
    Indeed, this is a direct consequence of th fact that,
    for any two distinct edges $(a, b)$ and $(a', b')$ in
    $\vv{\mathcal{M}}$ we have $\pi(a) < \pi(a')$ if and only if
    $\pi(b) < \pi(b')$.
    \qed
\end{proof}

@@

    An increasing pattern before and above a decreasing pattern.
        scale=0.2,
      % increasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,5.8) grid (5.2,11.2);
      \foreach \x/\y in {0/6,1/7,2/8,3/9,4/10,5/11} {
      % decreasing pattern
      \node [label] (a) at (1,4) {$a$};
      \node [label] (b) at (4,4) {$b$};
      \node [label] (ap) at (7,-2) {$a'$};
      \node [label] (bp) at (10,-2) {$b'$};
      \node (nu1) at (2.5,3) {$\sigma_1$};
      \node (nu2) at (8.5,-3) {$\sigma_2$};
      \draw [line width=1pt,->,>=latex']
      (1,7) .. controls +(0,12) and +(0,10) .. (7,4);
      \draw [line width=1pt,->,>=latex']
      (4,10) .. controls +(0,12) and +(0,10) .. (10,1);
    \;
        scale=0.2,
      % increasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,5.8) grid (5.2,11.2);
      \foreach \x/\y in {0/6,1/7,2/8,3/9,4/10,5/11} {
      % decreasing pattern
      \node [label] (a) at (1,4) {$a'$};
      \node [label] (b) at (4,4) {$b'$};
      \node [label] (ap) at (7,-2) {$a$};
      \node [label] (bp) at (10,-2) {$b$};
      \node (nu1) at (2.5,3) {$\sigma_1$};
      \node (nu2) at (8.5,-3) {$\sigma_2$};
      \draw [line width=1pt,<-,>=latex']
      (1,7) .. controls +(0,12) and +(0,10) .. (7,4);
      \draw [line width=1pt,<-,>=latex']
      (4,10) .. controls +(0,12) and +(0,10) .. (10,1);
    \label{subfig:increasing before and above decreasing}
    A decreasing pattern before and below an increasing pattern.
      scale=0.2,
    % decreasing pattern
    \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,-0.2) grid (5.2,5.2);
    \foreach \x/\y in {0/5,1/4,2/3,3/2,4/1,5/0} {
    % decreasing pattern
    \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,5.8) grid (11.2,11.2);
    \foreach \x/\y in {6/6,7/7,8/8,9/9,10/10,11/11} {
    \node [label] (a) at (1,-2) {$a$};
    \node [label] (b) at (4,-2) {$b$};
    \node [label] (ap) at (7,4) {$a'$};
    \node [label] (bp) at (10,4) {$b'$};
    \node (nu1) at (2.5,-3) {$\sigma_1$};
    \node (nu2) at (8.5,3) {$\sigma_2$};
    \draw [line width=1pt,->,>=latex']
    (1,4) .. controls +(0,10) and +(0,12) .. (7,7);
    \draw [line width=1pt,->,>=latex']
    (4,1) .. controls +(0,10) and +(0,12) .. (10,10);
    \;
        scale=0.2,
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,-0.2) grid (5.2,5.2);
      \foreach \x/\y in {0/5,1/4,2/3,3/2,4/1,5/0} {
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,5.8) grid (11.2,11.2);
      \foreach \x/\y in {6/6,7/7,8/8,9/9,10/10,11/11} {
      \node [label] (a) at (1,-2) {$a'$};
      \node [label] (b) at (4,-2) {$b'$};
      \node [label] (ap) at (7,4) {$a$};
      \node [label] (bp) at (10,4) {$b$};
      \node (nu1) at (2.5,-3) {$\sigma_1$};
      \node (nu2) at (8.5,3) {$\sigma_2$};
      \draw [line width=1pt,<-,>=latex']
      (1,4) .. controls +(0,10) and +(0,12) .. (7,7);
      \draw [line width=1pt,<-,>=latex']
      (4,1) .. controls +(0,10) and +(0,12) .. (10,10);
    \label{subfig:decreasing before and below increasing}
    Illustration of Lemma~\ref{lemma:at most one edge monotone}.

\begin{lemma}
  \label{lemma:at most one edge monotone}
  Let $\pi = \pi_1 \, \sigma_1 \, \pi_2 \, \sigma_2 \, \pi_3$
  be a permutation with $|\sigma_1| \geq 2$ and $|\sigma_2| \geq 2$,
  and $\vv{\mathcal{M}}$ be an oriented perfect matching of $G_\pi$
  that avoids
  \InclusionLL,~\InclusionLR,~\InclusionRL,~\InclusionRR,~\CrossingLR
  and \CrossingRL, and such that for any two distinct arcs $(a, a')$ and
  $(b, b')$ in $\vv{\mathcal{M}}$ we have $\pi(a) < \pi(b)$ if and only
  if $\pi(a') < \pi(b')$.
  The following assertions hold:
  \begin{enumerate}
    \item
    If $\sigma_1$ is increasing, $\sigma_2$ is decreasing, and
    $\sigma_1$ is above $\sigma_2$
    (see Fig.~\ref{subfig:increasing before and above decreasing}),
    then $\vv{\mathcal{M}}$ contains either at most one
    $(\sigma_1, \sigma_2)$-arc or at most one
    $(\sigma_2, \sigma_1)$-arc.
    \item
    If $\sigma_1$ is decreasing, $\sigma_2$ is increasing, and $\sigma_1$
    is below $\sigma_2$
    (see Fig.~\ref{subfig:decreasing before and below increasing}),
    then $\mathcal{M}$ contains at most one
    $(\sigma_1, \sigma_2)$-edge.
  \end{enumerate}
\end{lemma}

\begin{proof}[of Lemma~\ref{lemma:at most one edge monotone}]
  We only prove the first assertion since the proof for the second one
  is similar.
  Suppose, aiming at a contradiction, that the assertion does not hold.
  Therefore, since $\vv{\mathcal{M}}$ avoids
  \InclusionLL,~\InclusionLR,~\InclusionRL,~\InclusionRR,~\CrossingLR
  and \CrossingRL,
  $\vv{\mathcal{M}}$ contains either two crossing $(\sigma_1, \sigma_2)$-arcs
  or two crossing $(\sigma_2, \sigma_1)$-arcs.
  Suppose first that it contains two crossing $(\sigma_1, \sigma_2)$-arcs
  $(a, a')$ and $(b, b')$ with $a < b$.
  Therefore, we have $\pi(a) < \pi(b)$ (since $\sigma_1$ is increasing)
  and $\pi(a') > \pi(b')$ (since
  $\sigma_2$ is decreasing), which contradicts our hypothesis about
  $\vv{\mathcal{M}}$.
  If $\vv{\mathcal{M}}$ contains two crossing $(\sigma_2, \sigma_1)$-arcs
  $(a, a')$ and $(b, b')$ with $a < b$, then
  we have $\pi(a) > \pi(b)$ (since $\sigma_2$ is decreasing)
  and $\pi(a') < \pi(b')$ (since
  $\sigma_2$ is increasing), which contradicts our hypothesis about
  $\vv{\mathcal{M}}$.
  \qed
\end{proof}
